---
title: A tidystats Example
description: "I illustrate how to use my tidystats software to analyze and report the results of a replication study that was part of the Many Labs 1 project."
date: 2021-04-25
categories:
  - tidystats
  - statistics
  - tutorial
code-fold: true
bibliography: references.bib
toc: true
resources:
  - tidystats-example.json
  - tidystats-example.mp4
---

## Analyzing the Data

Lorge and Curtiss (1936) examined how a quotation is perceived when it is attributed to a liked or disliked individual. In one condition the quotation was attributed to Thomas Jefferson and in the other it was attributed to Vladimir Lenin. They found that people agree more with the quotation when the quotation was attributed to Jefferson than Lenin. In the Many Labs replication study [@klein2014], the quotation was attributed to either George Washington, the liked individual, or Osama Bin Laden, the disliked individual. The also used a different quotation, which was:

> I have sworn to only live free, even if I find bitter the taste of death.

We are again interested in testing whether the source of the quotation affects how it is evaluated. The evaluation was assessed on a 9-point Likert scale ranging from 1 (strongly agree) to 9 (strongly disagree). I reverse coded this in the data that we'll use. You can follow along by copy-pasting the code from this example.

Before getting into how `tidystats` should be used, let's first simply analyze the data. I have designed `tidystats` to be minimally invasive. In other words, to use `tidystats`, you do not need to substantially change your data analysis workflow.

We'll start with a basic setup where we load some packages and the data.

```{r}
#| label: setup
#| message: false

# Load necessary packages
library(tidyverse)
library(tidystats)
library(viridis)
library(knitr)

# Load example data
data <- data(quote_source)

# Set the default ggplot theme
theme_set(theme_minimal())

# Set options
options(
  knitr.kable.NA = "-",
  digits = 2
)
```

Our main effect of interest is the difference in responses to the quote between the two conditions. Here I visualize this different with a violin plot.

```{r}
#| label: fig-quote-source
#| fig-cap: Difference in responses between the two conditions
#| warning: false
ggplot(quote_source, aes(x = source, y = response, fill = source)) +
  geom_violin(width = .5, alpha = .85) +
  stat_summary(fun = "mean", geom = "point") +
  labs(x = "Quote source", y = "Quote agreement") +
  scale_y_continuous(breaks = c(1, 3, 5, 7, 9)) +
  scale_fill_viridis(option = "mako", discrete = TRUE, begin = .25, end = .75) +
  guides(fill = "none")
```

This looks like the effect is in the expected direction. Participants agreed more with the quotation when they believed the quote to be from George Washington compared to Osama Bin Laden.

Regarding descriptives, `tidystats` comes with its own functions to calculate descriptives. One of them is the `describe_data` function, inspired by the `describe` function from the `psych` package. You can use it together with `group_by` from the `dplyr` package to calculate a set of descriptives for multiple groups.

```{r}
#| label: tbl-response-descriptives
#| tbl-cap: Response descriptives
quote_source %>%
  group_by(source) %>%
  describe_data(response) %>%
  select(-var) %>%
  kable(col.names = c("Source", "Missing", "N", "M", "SD", "SE", "Min", "Max", 
    "Range", "Median", "Mode", "Skew", "Kurtosis")) 
```

To test whether the difference in agreement between the two sources is statistically significant, we perform a *t*-test. Normally, we would just run the t-test like so:

```{r}
#| label: t-test-normal
#| eval: false
t.test(response ~ source, data = quote_source)
```

However, since we want to use `tidystats` to later save the statistics from this test, we will store the output of the *t*-test in a variable. This, and the final section of R code, will be the only thing you need to change in order to incorporate `tidystats` in your workflow.

Once you've stored the result of the *t*-test in a variable, you can look at the output by sending it the console, which will print the output.

```{r}
#| label: t-test-tidystats
main_test <- t.test(response ~ source, data = quote_source)
main_test
```

This shows us that there is a statistically significant effect of the quote source, consistent with the hypothesis.

Next, let's run some additional analyses. One thing we can test is whether the effect is stronger in the US compared to non-US countries. To test this, we perform a regression analysis. Here we also store the result in a variable, but this is actually quite common in regression analyses because you want to apply the `summary` function to this variable in order to obtain the inferential statistics.

```{r}
#| label: lm-region
us_moderation_test <- lm(response ~ source * us_or_international, 
  data = quote_source)
summary(us_moderation_test)
```

There appears to be a significant interaction. Let's inspect the interaction with a graph:

```{r}
#| label: fig-interaction
#| fig-cap: Source by region interaction
#| warning: false
ggplot(quote_source, aes(x = us_or_international, y = response, 
    fill = source)) +
  geom_violin(alpha = .85) +
  stat_summary(fun.data = "mean_cl_boot", position = position_dodge(.9)) +
  labs(x = "Region", y = "Quote agreement", fill = "Source") +
  scale_y_continuous(breaks = c(1, 3, 5, 7, 9)) +
  scale_fill_viridis(option = "mako", discrete = TRUE, begin = .25, end = .75)
```

We see that the effect of the source appears to be larger in the US. Given that the positive source was George Washington, this makes sense.

Let's do one more analysis to see whether the effect is stronger in a lab setting compared to an online setting.

```{r}
#| label: lm-setting
lab_moderation_test <- lm(response ~ source * lab_or_online, data = quote_source)
summary(lab_moderation_test)
```

We see no significant interaction in this case. This means we do not find evidence that running the study in an online setting significantly weakens the effect; good to know!

## Applying tidystats

Now let's get to `tidystats`. We have three analyses we want to save: a *t*-test and two regression analyses. We stored each of these analyses in separate variables, called `main_test`, `us_moderation_test`, and `lab_moderation_test`.

The main idea is that we will add these three three variables to a list and then save the list as a file on our computer. You create an empty list using the `list` function. Once you have an empty list, you can add statistics to this list using the `add_stats` function. `add_stats` accepts a list as its first argument, followed by a variable containing a statistics model. In our case, this means we need to use the `add_stats` function three times, as we have three different analyses we want to save. Since this can get pretty repetitive, we will use the piping operator to pipe the three steps together and save some typing.

Before we do so, however, note that we can take this opportunity to add some meta-information to each test. For the sake of this example, let's say that the *t*-test was our primary test. We also had a suspicion that the location (US vs. international) would matter, but it wasn't our main interest. Nevertheless, we preregistered these two analyses. During data analysis, we figured that it might also matter whether the study was conducted in the lab or online, so we tested it. This means that this is an exploratory analysis. With `add_stats`, we can add this information when we add the test to our empty list.

In the end, the code looks like this:

```{r}
#| label: applying-tidystats
results <- list()

results <- results %>%
  add_stats(main_test, type = "primary", preregistered = TRUE) %>%
  add_stats(us_moderation_test, type = "secondary", preregistered = TRUE) %>%
  add_stats(lab_moderation_test, type = "secondary", preregistered = FALSE)
```

I recommend to do this at the end of the data analysis script in a section called 'tidystats'. This confines most of the `tidystats` code to a single section, keeping it organized, and it will keep most of your script readable to those unfamiliar with `tidystats`.

After all the analyses are added to the list, the list can be saved as a .json file to your computer's disk. This is done with the `write_stats` function. The function requires the list as its first argument, followed by a file path. I'm a big fan of using RStudio Project files so that you can define relative file paths. In this case, I create the .json file in the 'Data' folder of my project folder.

```{r}
#| label: write-stats
#| eval: false
write_stats(results, "tidystats-example.json")
```

If you want to see what this file looks like, you can inspect it [here](tidystats-example.json "tidystats JSON file example"). Open the file in a text editor to see how the statistics are structured. As you will see, it is not easy for our human eyes to quickly see the results, but it's easy for computers.

Once you've saved the file, you can share the file with others or use it to report report the results in your manuscript using the [Word add-in](https://www.tidystats.io/word-add-in.html "tidystats Word add-in").

That marks the end of this `tidystats` example. If you have any questions, please check out the [tidystats website](https://tidystats.io) or contact me via [Twitter](https://twitter.com/tidystats).

*This post was last updated on `r format(Sys.Date(), "%Y-%m-%d")`.*

## References
