---
title: "Sad wild animal facts"
description: "A collection of facts about wild animals that make me sad."
date: 2024-03-11
categories:
  - wild animals
  - wild animal suffering
toc: true
bibliography: references.bib
---

> Let us understand, once for all, that the ethical process of society depends, not on imitating the cosmic process, still less on running away from it, but in combating it.
>
> Thomas Huxley

This post serves to illustrate why I agree with the quote above. Below I list examples of what happens in nature that I think are just horrible and that I wish we could do something about.

The list is not complete and I intend to continuously update it when I encounter more sad facts.

*Last updated: `r lubridate::stamp_date("March 1, 1999", quiet = TRUE)(Sys.Date())`*

```{r}
#| label: setup
#| message: false
#| echo: false
data <- readr::read_csv("./data/animals.csv", na = c("", "-"))
data <- dplyr::filter(data, include == 1)

categories <- data |>
  dplyr::pull(category) |>
  unique()
```

```{r}
#| label: sad-animal-fact
#| results: asis
#| echo: false
for (current_category in categories) {
  subset <- dplyr::filter(data, category == current_category)
  cat(paste("##", current_category))
  cat("\n\n")
  for (i in 1:nrow(subset)) {
    name <- subset$name[i]
    latin_name <- subset$latin_name[i]
    description <- subset$description[i]
    citekey <- subset$citekey[i]

    # Find and replace the final period with the citekey and period
    description <- stringr::str_replace(
      description,
      pattern = ".$",
      replacement = paste0(" [@", citekey, "].")
    )

    name <- dplyr::if_else(is.na(name), latin_name, name)

    image <- stringr::str_to_lower(name)
    image <- stringr::str_replace_all(image, " ", "-")
    image <- paste0("images/", image, ".jpg")

    cat(paste("###", name))
    cat("\n\n")

    if (file.exists(image)) {
      cat(paste0("![](", image), ")")
      cat("\n\n")
    }

    cat(paste("**Scientific name**:", latin_name))
    cat("\n\n")

    cat(description)
    cat("\n\n")
  }
}
```