---
title: Understanding regression (part 1)
description: This is Part 1 of a series on understanding regression. Using simulated data, we explore where the intercept estimate comes from in a simple regression model and discover why it equals the mean when we minimize squared residuals.
date: 2020-08-01
updated: 2025-01-10
tags:
  - statistics
  - tutorial
  - regression
code-fold: true
toc: true
---

Statistical regression techniques are an important tool in data analysis. As a behavioral scientist, I use it to test hypotheses by comparing differences between groups of people or testing relationships between variables. While it is easy to run regression analyses using popular software tools, like SPSS or R, it often remains a black box that is not well understood. In fact, I do not believe I actually understand regression. Not fully understanding the mechanics of regression could be okay, though. After all, you also don't need to know exactly how car engines work in order to drive a car. However, I think many users of regression have isolated themselves too much from the mechanics of regression. This may be a source of some mistakes researchers make, such as including every available variable into a model without understanding what the regression is actually doing. If you're using regression to try and make inferences about the world, it's probably a good idea to know what you're doing.

In this series of posts, I'll work through the mechanics of regression to build a genuine intuition for how it works. This is Part 1.

Feel free to follow me along by copy-pasting the code from each step.

## Setup

To figure out regression, we need data. For this tutorial, I'll use simulated data that could represent many different scenarios---test scores (like IQ or standardized tests), measurements of a physical trait (like height or reaction time), or ratings on a psychological scale. The key is that the data follows a pattern commonly seen in nature and social sciences.

In case you're following along, start by loading the required packages and a custom function to calculate the mode. You can also copy the styling options I use, such as my ggplot2 defaults.

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(here)
library(gt)

# A custom function to calculate the mode
mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Theme settings
theme_set(theme_minimal(base_size = 16))
primary <- "#16a34a"

# Set seed for reproducibility
set.seed(42)

# Generate simulated data (mean = 100, sd = 15, like IQ scores)
n <- 25
scores <- rnorm(n, mean = 100, sd = 15) |> round(1)
data <- tibble(
  id = 1:n,
  score = scores
)
```

Let's visualize our data. We have 25 observations, and you can think of these as scores on some measure. Below I've plotted each observation.

```{r}
#| label: fig-data
#| fig-width: 10
ggplot(data, aes(x = id, y = score)) +
  geom_bar(stat = "identity", fill = primary) +
  labs(x = "Observation", y = "Score") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = seq(0, 25, by = 5)) +
  theme(
    panel.grid.major.x = element_blank()
  )
```

We see that the lowest score is `r min(data$score)` and the highest is `r max(data$score)`. The average score is `r round(mean(data$score), 1)`.

## The simplest model

In order to understand these scores, we need to come up with a statistical model. In a way, this can be considered a description problem. How can we best describe the different scores that we have observed? The simplest description is a single number. We can say that all observations have a score of, say, 90. In other words:

`score = 90`

Of course, this is just one among many possible models. Below I plot three different models, including our `score = 90` model.

```{r}
#| label: fig-three-models
#| fig-width: 11
#| fig-height: 6
ggplot(data, aes(x = id, y = score)) +
  geom_bar(stat = "identity", fill = primary) +
  geom_abline(intercept = c(90, 100, 110), slope = 0, linetype = "dashed") +
  annotate(
    geom = "label",
    label = c("score = 90", "score = 100", "score = 110"),
    x = 24,
    y = c(92, 102, 112),
    hjust = 1,
    vjust = 0,
    size = 5
  ) +
  labs(x = "Observation", y = "Score", linetype = "") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = seq(0, 25, by = 5)) +
  scale_linetype_manual(values = c("dashed", "dotted", "dotdash")) +
  coord_cartesian(clip = "off") +
  theme(
    panel.grid.major.x = element_blank()
  )
```

While a model like `score = 90` is a valid model, it is not a very good model. It may perfectly describe one or two observations, but it inaccurately describes most of the others. The other models might be better or worse, depending on the actual data. How do we decide which model is the better model? In order to answer that question, we need to consider the model's error.

## The error of models

The error of a model is the degree to which the model inaccurately describes the data. There are several ways to calculate that error, depending on how you define error. We will cover three of them.

The first definition of error is simply the number of times the model inaccurately describes the data. For each observation we check whether the model correctly describes it or not. We then sum the number of misses and consider that the amount of error for that model. For our `score = 90` model, we would count how many observations have a score different from 90.

We can now compare different models to one another by calculating the error for a range of models. Below I plot the number of errors for many different models, starting with `score = 70`, up to `score = 130`, in steps of 0.1.

```{r}
#| label: fig-error-sum
errors_binary <- expand_grid(
  model = seq(from = 70, to = 130, by = 0.1),
  score = data$score
) |>
  mutate(error = if_else(abs(score - model) == 0, 0, 1)) |>
  group_by(model) |>
  summarize(error_sum = sum(error))

ggplot(errors_binary, aes(x = model, y = error_sum)) +
  geom_line(color = primary, linewidth = 1.25) +
  coord_cartesian(ylim = c(0, 25)) +
  labs(x = "Model (score = x)", y = "Error (sum of errors)")
```

We see that almost all models perform poorly. The number of errors range from `r min(pull(errors_binary, error_sum))` to `r max(pull(errors_binary, error_sum))` and most models have a sum of `r mode(pull(errors_binary, error_sum))` errors, which means they do not accurately describe any of the observations. The best models might correctly describe one or two observations, but that still leaves 23 or 24 errors.

Despite there being a best model, it's still a pretty poor model. After all, it is wrong most of the time. Perhaps there are some models that outperform this model, but it's unlikely. That's because we're defining error here in a very crude way. The model needs to exactly match the observed score, or else it counts as an error. Saying a score is 90, while it is in fact 95, is as wrong as saying the score is 50.

Instead of defining error in this way, we can redefine it so that it takes into account the *degree* of error. We can define error as the difference between the actual data point and the model's value. So, in the case of our `score = 90` model, an actual score of 95 would have an error of 95 - 90 = 5. This definition of error is often referred to as the residual.

Below I plot the residuals for our `score = 90` model.

```{r}
#| label: fig-residuals
#| fig-width: 10
ggplot(data, aes(x = id, y = score)) +
  geom_bar(stat = "identity", fill = primary) +
  geom_segment(aes(xend = id, y = 90, yend = score), linetype = 2) +
  geom_point() +
  geom_abline(intercept = 90, slope = 0) +
  labs(x = "Observation", y = "Score") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = seq(0, 25, by = 5)) +
  theme(
    panel.grid.major.x = element_blank()
  )
```

We can add up all of the (absolute) residuals to determine the model's error. Just like with the binary definition of error, we can then compare multiple models. This is what you see in the graph below. For each model, from `score = 70` to `score = 130`, we calculated the absolute residuals and added them together.

```{r}
#| label: fig-error-residuals
errors_residuals <- expand_grid(
  model = seq(from = 70, to = 130, by = 0.1),
  score = data$score
) |>
  mutate(error = abs(score - model)) |>
  group_by(model) |>
  summarize(error_sum = sum(error))

ggplot(errors_residuals, aes(x = model, y = error_sum)) +
  geom_line(color = primary, linewidth = 1.25) +
  labs(x = "Model (score = x)", y = "Error (sum of residuals)")
```

This graph looks very different compared to the graph where we calculated the error defined as the sum of misses. Now we see that a minimum appears. Unlike the binary definition of error, it looks like there are fewer *best* models. More importantly, though, we have defined error in a less crude manner, meaning that the better models indeed capture the data much better than before.

But we might still not be entirely happy with this new definition of error either. Calculating the sum of absolute residuals for each model comes with another conceptual problem.

When you sum the absolute errors, four errors of 1 are equal to a single error of 4. In other words, you could have a model that is slightly off multiple times or one that might make fewer, but larger, errors. Both would be counted as equally wrong. That seems problematic. Conceptually speaking, we might find it more problematic when a model is very wrong than when the model is slightly off multiple times. If we think that, we need another definition of error.

To address this issue, we can square the residuals before adding them together. That way, larger errors become relatively larger compared to smaller errors. Using our previous example, summing four residuals of 1 remains 4, but a single residual of 4 becomes 4² = 16. The model now gets punished more severely for making large mistakes.

Using this new definition of error, we again plot the error for each model, from 70 to 130.

```{r}
#| label: fig-error-squared-residuals
errors_squared_residuals <- expand_grid(
  model = seq(from = 70, to = 130, by = 0.1),
  score = data$score
) |>
  mutate(error = abs(score - model)^2) |>
  group_by(model) |>
  summarize(error_sum = sum(error))

ggplot(errors_squared_residuals, aes(x = model, y = error_sum)) +
  geom_line(color = primary, linewidth = 1.25) +
  geom_vline(xintercept = mean(data$score), linetype = 2) +
  labs(x = "Model (score = x)", y = "Error (sum of squared residuals)")
```

We see a smooth curve, with a clear minimum indicated by the vertical dashed line. This vertical line indicates the model that best describes the data. What is the value of the best model exactly? In this case, the answer is `r round(mean(data$score), 1)`. And it turns out, there is an easy way to determine this value.

## The data-driven model

Rather than setting a specific value and seeing how it fits the data, we can also use the data to determine the best-fitting value. In the previous graph we saw that the best fitting model is one where the score is equal to `r round(mean(data$score), 1)`. This value turns out to be the mean of the different scores we have observed in our sample. Had we defined error as simply the sum of absolute residuals, this would be a different value. In fact, the best fitting value would then be equal to `r round(median(data$score), 1)`, or the median. And had we used the binary definition of error, the best fitting value would be the mode, which in our case is: `r mode(data$score)`.

Note that there is not always a unique answer to which model is the best fitting model, depending on the error definition. For example, it is possible that there are multiple modes. If you use the binary definition of error, that would mean there are multiple equally plausible models. This can be another argument to not define a model's error in such a crude way.

The table below shows an overview of which technique can be used to find the best fitting value, depending on the error definition.

```{r}
#| label: tbl-errors
tibble(
  error_definition = c(
    "sum of errors",
    "sum of absolute residuals",
    "sum of squared residuals"
  ),
  estimation_technique = c("mode", "median", "mean")
) |>
  gt() |>
  cols_label(
    error_definition = "Error definition",
    estimation_technique = "Estimation technique"
  )
```

We can now update our model to refer to the estimation technique, rather than a fixed value. Given that the third definition of error seems to be most suitable, both pragmatically and conceptually, we'll use the mean:

`score = mean(score)`

This is also the value you get when you perform a regression analysis in R.

```{r}
#| label: lm-score-intercept-model
lm(score ~ 1, data = data)
```

By regressing score onto 1 we are telling R to run an intercept-only model. This means that R will estimate which value will best fit all the values in the outcome variable, just like we have done ourselves earlier by testing different models such as `score = 90`.

The result is an intercept value of `r round(mean(data$score), 1)`, which matches the mean of the scores.

## Conclusion

So, we now know where the intercept comes from when we run an intercept-only model: it is the mean of the data we are trying to model.

Note that it is the mean because we defined the model's error as the sum of *squared* residuals. Had we defined the error differently, the intercept would be different. With the sum of *absolute* residuals, the intercept would be the median. With the sum of errors, it would be the mode.

Why did we use the sum of squared residuals? We had a conceptual reason: we wanted to punish larger residuals relatively more than several smaller errors. But it turns out there is another reason to favor squared residuals, which has to do with a nice property of the mean compared to the median. This will be covered in Part 2 of 'Understanding Regression'.

## References

- Fox, J., & Weisberg, S. (2019). *An R Companion to Applied Regression* (3rd ed.). Sage.

- James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning with Applications in R* (2nd ed.). Springer.