---
title: Replacing standard tests with SEM
description: "Description"
date: 2022-06-14
categories:
  - statistics
  - tutorial
code-fold: true
toc: true
draft: true
---

```{r}
#| label: setup
#| message: false

# Load packages
library(tidyverse)
library(lavaan)

# Set options
options(
  knitr.kable.NA = "-", 
  digits = 2
)

theme_set(theme_minimal())
```

In this post I will show how to use structural equation modeling to run common statistical tests. Although this will generally involve more code and a steeper learning curve to running these tests, it will promote a more general understanding of these tests and also enable you to go beyond the usual norms and run better analyses.

## T-test

Let's begin with a simple *t*-test. Below is the code to run a *t*-test using what is probably the most common function for this test in R.

```{r}
#| label: t-test
t.test(extra ~ group, data = sleep)
lm(extra ~ group, data = sleep) |> summary()


sleep2 <- mutate(sleep, group = as.numeric(group))
```

```{r}
#| label: lavaan-t-test
model <- "
  # Regression model
  extra ~ 1 + group

  group ~~ group
"

fit <- sem(model = model, data = sleep2)
summary(fit)
```
