---
title: "Understanding Regression (Part 1): Getting Started"
description: "This is the first post in a series on understanding regression. In this first post I focus on what the main question is that we should be asking when using regression."
date: 2025-04-24
categories:
  - statistics
  - tutorial
  - regression
code-fold: true
toc: true
---

```{r}
#| label: setup
#| message: false
library(tidyverse)

# Theme settings
primary <- "#16a34a"

theme_set(theme_minimal(base_size = 14))
update_geom_defaults(
  "histogram",
  aes(fill = primary, color = "white")
)
```

## Introduction

I'm a behavioral scientist with several scientific publications to my name in which I've used statistical tools like regression. Yet, despite my experience with this statistical technique, I don't feel like I *really* understand regression. I know *how* to use it—I can run models and report the results-but there are times when I'm just running code because I've been told that's how you do it. When things get complicated, and sometimes not even that complicated, I find myself relying almost entirely on conventions rather than on my own understanding.

You might feel the same way. Regression is often taught as a black box: you run the code or click the buttons and copy the output. The focus in many courses is on performing statistics rather than on building a conceptual understanding. I've also noticed a lot of statistical teaching takes a more mechanical approach. You're given formulas to memorize or asked to calculate statistics by hand. I can see the appeal of this approach, since working through the steps yourself can build intuition, but formulas rarely help me understand something, and calculating things by hand only takes you so far.

What I need, and what I suspect many others need, is an approach that is more about creating a *conceptual* understanding that makes regression make sense. That's what this series is about.

## An example

To build this conceptual understanding, I'll work through examples using real data. I'll use data from Richard McElreath's excellent book *Statistical Rethinking*. The data is a partial census of the !Kung San people, compiled from interviews conducted by Nancy Howell in the late 1960s. We'll focus on heights of adults (18 years or older).

Here's what the first few rows of the data look like:

```{r}
#| label: data
data <- read_csv("Howell1.csv")
data <- filter(data, age >= 18)

head(data)
```

And here's a histogram of all the heights:

```{r}
#| label: histogram
ggplot(data, aes(x = height)) +
  geom_histogram()
```

Let's immediately run a regression so we have some concrete numbers to look at. The simplest model we can run is one in which we regress heights onto... nothing; this is called an **intercept-only model**.

```{r}
#| label: model
#| code-fold: show
model <- lm(height ~ 1, data = data)
```

We use `summary()` to get the numbers we need.

```{r}
#| label: model-summary
summary(model)
```

As you can see in the output, we get an estimate of `r round(coef(model)[[1]], 2)` cm and a standard error of `r round(summary(model)$coefficients[, "Std. Error"], 2)` cm.

Here's where I want to pause and ask ourselves the question, why do we actually use these statistical tools? Why do we use regression?

## Why use regression?

When we run a regression model, we're typically interested in one or more of the following goals:

1.  **Estimation**: What are the model parameters (like the mean) and how uncertain are we about them?
2.  **Testing**: Is a parameter compatible with some reference value (like zero)?
3.  **Prediction**: What would we expect to observe in new data?

In the behavioral science literature I'm familiar with, testing is by far the most common goal. Researchers want to know whether an effect is "significant", which means running a model and checking whether the estimate is compatible with zero. Estimation is also common, with researchers reporting coefficients, standard errors, and confidence intervals. Prediction is less popular (authors rarely report prediction intervals), but is also important.

I'm looking for a way of thinking that makes it easier to understand how regression can be used for each of those goals and I think it comes down to the following question:

**What distribution might have generated this data?**

Instead of starting with a procedure — run a model, get an estimate — you start by imagining how the data came to be. Some process generated these numbers. That process had some underlying shape: values that were more likely, values that were less likely. A distribution is a way of describing that shape formally. You may already be familiar with some: the normal distribution, the Poisson distribution, or perhaps the binomial distribution.

For our height data, we can propose that heights follow a normal distribution. This is a bell-shaped curve defined by two parameters: μ (the mean) and σ (the standard deviation).

These two parameters give us everything we need for estimation, testing, and prediction:

-   **For estimation**: μ tells us the typical height, and we can quantify how uncertain we are about that estimate
-   **For testing**: Once we know the estimate and its uncertainty, we can ask whether it's compatible with a specific value (like zero)
-   **For prediction**: Both μ and σ together describe the full distribution, which we can use to predict what heights we'd expect in new individuals

Here are the heights again, but now with a normal distribution on top of the histogram.

```{r}
#| label: distribution
#| fig-cap: "Heights with normal distribution overlay"
sample_mean <- mean(data$height)
sample_sd <- sd(data$height)

ggplot(data, aes(x = height)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 5) +
  stat_function(
    fun = dnorm,
    args = list(mean = sample_mean, sd = sample_sd),
    linewidth = 1,
    color = "black",
    linetype = "dashed"
  ) +
  labs(
    x = "Height (cm)",
    y = "Density"
  )
```

The dashed line is the normal distribution we're proposing heights are drawn from.

Our `lm()` model estimated the parameters that define the distribution. The intercept (`r round(coef(model)[[1]], 2)`) is our estimate of μ, and the residual standard error is our estimate of σ.

## What's ahead

In this series, I'm going to build up an understanding of regression from this foundation. The core question will always be: **What distribution might have generated this data?**

Here are the main steps we'll tackle:

1.  **Choose a distribution**: What distribution might describe the data?
2.  **Estimate parameters**: How do we estimate its parameters from our data?
3.  **Quantify uncertainty**: How certain can we be about those estimates?
4.  **Test hypotheses**: Are the estimates compatible with specific values?
5.  **Make predictions**: What would we expect to observe in new data?
6.  **Add predictors**: What happens when the distribution's parameters depend on other variables (like height depending on sex or age)?

Each post in the series will tackle one or more of these steps, building up our understanding gradually.

## Summary

This post introduces the core perspective to hopefully have regression make sense: **regression is about choosing and fitting distributions to data.**

When you run a regression model, you're proposing that your data follows a specific distribution and you're estimating that distribution's parameters. Those parameters let you estimate typical values, quantify uncertainty, test hypotheses, and predict new observations.

In the next post, we'll dig deeper into the normal distribution and why it's often a sensible default.