---
title: "Understanding Regression (Part 1): Getting Started"
description: "This is the first post in a series on understanding regression. In this first post I focus on what the main question is that we should be asking when we think we should use regression."
date: 2025-01-15
categories:
  - statistics
  - tutorial
  - regression
code-fold: true
toc: true
draft: true
---

```{r}
#| label: setup
#| message: false
library(tidyverse)

# Theme settings
primary <- "#16a34a"

theme_set(theme_minimal(base_size = 16))
update_geom_defaults(
  "histogram",
  aes(fill = primary, color = "white")
)
```

## Introduction

This is the first post of a series of blog posts about understanding regression.

I'm a behavioral scientist with several scientific publications to my name in which I've used statistical tools like regression to answer questions. Yet, despite my experience, I admittedly don't feel like I *really* understand regression. I know *how* to use it—I can run models and report the results. But there are times when I'm just running code because I've been told that's how you do it. When things get complicated, and sometimes not even that complicated, I find myself relying almost entirely on conventions rather than understanding it myself.

You might feel the same way. Regression is often taught as a black box: you run the code or click the buttons and copy the output. The focus is on performing statistics rather than on building a conceptual understanding.

I also sometimes see a focus on the mechanics of statistics in classes on regression. For example, you might be presented with formulas or you're asked to manually calculate some statistic. While I'm sure formula's and manually calculating statistics can help some people understand statistics, but they don't really work for me. Formulas are rarely helpful for me to understand something (in fact, they often hinder it) and manually calculating some statistics only takes you so far.

What I need, and what I suspect many others need, is an approach that is more about creating a *conceptual* understanding that makes regression make sense. That's what this series is about.

## An example

To build this conceptual understanding, I'll work through examples using real data. I'll use data from Richard McElreath's excellent book *Statistical Rethinking*, which also inspired the approach I'll be taking. The data is a partial census of the !Kung San people, compiled from interviews conducted by Nancy Howell in the late 1960s. We'll focus on heights of adults (18 years or older).

Here's what the first few rows of the data look like:

```{r}
#| label: data
data <- read_csv("Howell1.csv")
data <- filter(data, age >= 18)

head(data)
```

And here's what all the heights look like, as a histogram:

```{r}
#| label: histogram
ggplot(data, aes(x = height)) +
  geom_histogram()
```

Let's immediately run a regression so we have some numbers to look at and ask ourselves what they mean. The simplest model we can run is one in which we regress heights onto... nothing; this is called an **intercept-only model**.

```{r}
#| label: model
model <- lm(height ~ 1, data = data)

summary(model)
```

As you can see in the output produced by running the `summary()` function, we get an estimate of `r round(coef(model)[[1]], 2)` cm and a standard error of `r round(summary(model)$coefficients[, "Std. Error"], 2)` cm.

Here's where I want to pause and ask some questions:

-   What do these numbers *mean*? The estimate is `r round(coef(model)[[1]], 2)` cm, but what is that estimating?
-   Why do they have these particular values? The estimate turns out to be the mean of the sample, but why is it the mean? Why not the median or some other number?
-   Why do we want these specific numbers?

## Why use regression?

When we run a regression model, we're typically interested in one or more of the following goals:

1.  **Estimation**: What are the model parameters (like the mean) and how uncertain are we about them?
2.  **Testing**: Is a parameter compatible with some reference value (like zero)?
3.  **Prediction**: What would we expect to observe in new data?

In the behavioral science literature I'm familiar with, testing is by far the most common goal. Researchers want to know whether an effect is "significant," which means running a model and checking whether the estimate is compatible with zero. Estimation is also common, with researchers reporting coefficients, standard errors, and confidence intervals. Prediction is less popular (authors rarely report prediction intervals, for instance), but is also important.

I think all three goals are often what's of interest, so I want to take a perspective that encompasses all of them.

My current thinking is that all three goals can best be understood through a single unifying question:

**What distribution might have generated this data?**

For our height data, we can propose that heights follow a normal distribution—a bell-shaped curve defined by two parameters: μ (the mean) and σ (the standard deviation).

These two parameters give us everything we need for estimation, testing, and prediction:

-   **For estimation**: μ tells us the typical height, and we can quantify how uncertain we are about that estimate
-   **For testing**: Once we know the estimate and its uncertainty, we can ask whether it's compatible with a specific value (like zero)
-   **For prediction**: Both μ and σ together describe the full distribution, which we can use to predict what heights we'd expect in new individuals

Here are the heights again, but now with a normal distribution on top of the data points:

```{r}
#| label: distribution
#| fig-cap: "Heights with normal distribution overlay"

sample_mean <- mean(data$height)
sample_sd <- sd(data$height)

ggplot(data, aes(x = height)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 5) +
  stat_function(
    fun = dnorm,
    args = list(mean = sample_mean, sd = sample_sd),
    linewidth = 1,
    color = "black",
    linetype = "dashed"
  ) +
  labs(
    x = "Height (cm)",
    y = "Density"
  )
```

The dashed line represents the normal distribution. We're proposing that heights are drawn from this distribution.

Our `lm()` model estimated the parameters that define the distribution. The intercept (`r round(coef(model)[[1]], 2)`) is our estimate of μ, and the residual standard error is our estimate of σ.

## What's ahead

In this series, I'm going to build up an understanding of regression from this foundation. The core question will always be: **What distribution might have generated this data?**

Here are the main steps we'll tackle:

1.  **Choose a distribution**: What distribution might describe the data?
2.  **Estimate parameters**: How do we estimate its parameters from our data?
3.  **Quantify uncertainty**: How certain can we be about those estimates?
4.  **Test hypotheses**: Are the estimates compatible with specific values?
5.  **Make predictions**: What would we expect to observe in new data?
6.  **Add predictors**: What happens when the distribution's parameters depend on other variables (like height depending on sex or age)?

Each post in the series will tackle one or more of these steps, building up our understanding gradually.

## Summary

This post introduces the core perspective that will guide this entire series: **regression is about choosing and fitting distributions to data.**

When you run a regression model, you're proposing that your data follows a specific distribution and you're estimating that distribution's parameters. Those parameters let you estimate typical values, quantify uncertainty, test hypotheses, and predict new observations.

In the next post, we'll dig deeper into the normal distribution and why it's often a reasonable distribution to use.