---
title: "Understanding Regression (Part 3): Estimating Parameters"
description: "We've chosen the normal distribution to describe heights. Now we need to estimate its parameters μ and σ from our sample. This post shows that lm() does exactly this: the intercept estimates μ (the mean) and the residual standard error estimates σ."
date: 2025-01-17
categories:
  - statistics
  - tutorial
  - regression
code-fold: true
toc: true
draft: true
---

```{r}
#| label: setup
#| message: false
library(tidyverse)

# Theme settings
primary <- "#16a34a"

theme_set(theme_minimal(base_size = 16))
update_geom_defaults(
  "histogram",
  aes(fill = primary, color = "white")
)
```

## Recap

In Part 2, I argued that the normal distribution is often the correct distribution for outcomes like heights, which are shaped by many small additive effects. It's also parsimonious: defined by just two parameters, it captures a center and some spread without adding claims about skewness, multiple groups, or hard boundaries.

Now we need to estimate those two parameters from our data.

## Estimating the parameters

The normal distribution is defined by two parameters:

- **μ (mu)**: the mean, where the center of the distribution is
- **σ (sigma)**: the standard deviation, how spread out the distribution is

If we can estimate these from our sample, we've characterized the entire distribution. The sample mean estimates μ and the sample standard deviation estimates σ. Let me calculate these using the same height data we've been working with:

```{r}
#| label: sample-stats
data <- read_csv("Howell1.csv")
data <- filter(data, age >= 18)

sample_mean <- mean(data$height)
sample_sd <- sd(data$height)
```

- **Sample mean**: `r round(sample_mean, 2)` cm
- **Sample SD**: `r round(sample_sd, 2)` cm

These are our estimates of μ and σ.

## What lm() is doing

Now let's run the same regression model from Part 1:

```{r}
#| label: model
model <- lm(height ~ 1, data = data)
summary(model)
```

Look at the output:

- The **intercept** is `r round(coef(model)[[1]], 2)` cm
- The **residual standard error** is `r round(sigma(model), 2)` cm

Compare these to what we calculated:

- Intercept (`r round(coef(model)[[1]], 2)`) = Sample mean (`r round(sample_mean, 2)`)
- Residual SE (`r round(sigma(model), 2)`) = Sample SD (`r round(sample_sd, 2)`)

They're identical. When there are no predictors, the residual standard error is the same as the sample standard deviation.

The name "residual standard error" hints at what's going on. A **residual** is the difference between an observed value and the model's estimate. For our intercept-only model, the estimate is the mean, so each residual is just a height minus the mean. The residual standard error then squares those residuals, sums them, divides by n − 1, and takes the square root. But that's exactly how the sample standard deviation is calculated: take each value, subtract the mean, square the differences, sum them, divide by n − 1, and take the square root. They're the same computation.

So `lm()` isn't doing anything exotic. The intercept is the sample mean (our estimate of μ) and the residual standard error is the sample standard deviation (our estimate of σ). 

## Visualizing the estimated distribution

Let me overlay the distribution with our estimated parameters on the histogram:

```{r}
#| label: fig-estimated-distribution
#| fig-cap: "Heights with normal distribution using estimated parameters"

ggplot(data, aes(x = height)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 5) +
  stat_function(
    fun = dnorm,
    args = list(mean = sample_mean, sd = sample_sd),
    linewidth = 1,
    color = "black",
    linetype = "dashed"
  ) +
  labs(
    x = "Height (cm)",
    y = "Density"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)))
```

This is our regression model: heights follow a normal distribution with mean `r round(sample_mean, 1)` cm and standard deviation `r round(sample_sd, 1)` cm.

## Summary

We started with a question: what distribution might have generated our height data? We proposed the normal distribution, and now we've estimated its parameters from our sample: μ ≈ `r round(sample_mean, 1)` cm and σ ≈ `r round(sample_sd, 1)` cm.

When we run `lm(height ~ 1)`, we're estimating these same parameters. The regression output gives us both values, along with additional information (like uncertainty about our estimates, which we'll explore later).

But these estimates are based on just one sample of 352 people. If we'd measured a different group, we'd get different numbers. How much would they vary? How certain can we be about our estimates? That's what we'll tackle in Part 4.
