---
title: "Understanding Regression (Part 7): The p-Value"
description: "We have a t-value and know it follows a t-distribution. In this post, we introduce hypothesis testing and the null hypothesis, and show how the t-distribution serves as a reference for evaluating whether our estimate is compatible with a specific claim — that probability is the p-value."
date: 2026-04-28
categories:
  - statistics
  - tutorial
  - regression
code-fold: true
toc: true
draft: true
---

```{r}
#| label: setup
#| message: false
library(tidyverse)

primary <- "#16a34a"

theme_set(theme_minimal(base_size = 16))
update_geom_defaults(
  "histogram",
  aes(fill = primary, color = "white")
)
```

## Recap

Part 6 introduced the t-value: the distance between our estimate and a reference value, expressed in units of the SE. We also saw that t-values follow the t-distribution rather than the normal, because using an estimated σ adds variability to the SE and produces heavier tails. In this post, we use that distribution to make sense of the last column in `lm()` output — `Pr(>|t|)` — and explain what that number means.

```{r}
#| label: data
data <- read_csv("Howell1.csv")
data <- filter(data, age >= 18)

sample_mean <- mean(data$height)
sample_sd <- sd(data$height)
n <- nrow(data)
se <- sample_sd / sqrt(n)

t_vs_155 <- (sample_mean - 155) / se
t_vs_0 <- sample_mean / se
```

## The logic of the p-value

Part 6 left us with a t-value: a number expressing how far our estimate sits from a reference value in units of the SE. [insert we also covered its distribution, the t-distribution]. But how do we go from here to something called the p-value? 

We want to be able to say that our estimate is some kind of meaningful value. That it is a signal, not noise. Usually we want to assert that the estimate we observed is not just the result of random sampling (i.e., is larger than 0) or that it is larger than some kind of value. The former is by far the most popular.

Our logic here becomes that in order to assert that our estimate is some kind of meaningful value, we will assume that some other model is true

The question is about compatibility. We have an estimate of μ — a value computed from our data — and we want to evaluate whether it is consistent with a specific claim. Is a mean height of `r round(sample_mean, 2)` cm consistent with the claim that the mean is 155 cm? Is it consistent with the claim that the mean is 0? These are different questions with different answers, and each one is a **hypothesis test**.

A hypothesis is a specific claim about a parameter. "The mean height is 155 cm" is a hypothesis. "This predictor has no effect" is a hypothesis. What makes hypothesis testing useful is that it gives us a principled way to assess whether our data are consistent with a given claim — not just a vague sense of closeness, but a quantified statement.

The specific claim we are evaluating is called the **null hypothesis**. The word "null" reflects that these claims often represent a baseline or default — something like "no effect" or "no difference." But the null hypothesis can be any specific value. What matters is that it is precise: a single number for μ.

Precision matters because it lets us go further than just using the reference value as a point to subtract from. If we assume the null hypothesis — that μ is 155 cm — our model now has both parameters specified: μ set to the null value and σ as our estimated standard deviation. With both parameters fixed, we already know from Part 6 what t-values look like: they follow the t-distribution. And because the null value is already subtracted in the numerator of t, the distribution is centered at 0 regardless of what the null value is. That is the **null distribution** — it is simply the t-distribution applied to the null model.

With the null distribution, we can ask where our observed t-value falls within it. If it sits near the center, our estimate is consistent with the null model. If it falls far in the tail, the null model does a poor job of explaining what we observed, and we have reason to doubt it. The proportion of the null distribution that falls at least as far from zero as our t-value is the **p-value**.

To see this in action, consider testing whether the mean height equals 155 cm. Our estimated mean is `r round(sample_mean, 2)` cm, giving:

$$t = \frac{`r round(sample_mean, 2)` - 155}{`r round(se, 2)`} \approx `r round(t_vs_155, 2)`$$

We can verify this by simulation — drawing samples with μ = 155 cm and our estimated standard deviation, computing a t-value from each, and checking whether the result matches the t-distribution.

```{r}
#| label: fig-null-distribution
#| fig-cap: "Distribution of t-values from 10,000 simulated samples where μ = 155 cm. The dashed curve is the t-distribution; the vertical line marks our observed t-value. The shaded regions show the proportion of t-values at least as extreme as ours."
set.seed(42)

n_simulations <- 10000

null_t_values <- replicate(n_simulations, {
  sim <- rnorm(n, mean = 155, sd = sample_sd)
  (mean(sim) - 155) / (sd(sim) / sqrt(n))
})

x_range <- seq(-5, 5, length.out = 1000)
t_curve <- tibble(x = x_range, y = dt(x_range, df = n - 1))

shade_left <- t_curve |> filter(x <= -abs(t_vs_155))
shade_right <- t_curve |> filter(x >= abs(t_vs_155))

p_value_sim <- mean(abs(null_t_values) >= abs(t_vs_155))
p_value_exact <- 2 * pt(abs(t_vs_155), df = n - 1, lower.tail = FALSE)

ggplot() +
  geom_histogram(
    data = tibble(t = null_t_values),
    aes(x = t, y = after_stat(density)),
    binwidth = 0.1
  ) +
  geom_area(
    data = shade_left,
    aes(x = x, y = y),
    fill = "black",
    alpha = 0.3
  ) +
  geom_area(
    data = shade_right,
    aes(x = x, y = y),
    fill = "black",
    alpha = 0.3
  ) +
  geom_line(
    data = t_curve,
    aes(x = x, y = y),
    linetype = "dashed",
    linewidth = 0.8
  ) +
  geom_vline(
    xintercept = t_vs_155,
    linewidth = 1
  ) +
  labs(x = "t value", y = "Density") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)))
```

The simulation confirms it: the histogram of t-values from the null model follows the t-distribution closely. The vertical line marks our observed t-value of `r round(t_vs_155, 2)`, which sits well within the bulk of the distribution. The shaded regions mark the proportion falling at least as far from zero in both directions.

In our simulation, `r round(p_value_sim * 100, 1)`% of t-values fell at least as far from zero as ours. Computing the same probability analytically from the t-distribution formula gives p = `r round(p_value_exact, 3)`. We fail to reject the null: our estimate is consistent with the null model.

Getting the interpretation right, though, requires thinking carefully about what exactly we're claiming. It's subtler than it first appears.

A few things are worth spelling out carefully.

**"At least as extreme" rather than "exactly this value."** A continuous distribution assigns probability zero to any single point, so asking "how likely is t = 1.47 exactly?" isn't a useful question. Instead we ask about tail areas: what proportion of the null distribution falls as far from zero as our t-value, or farther? That's the shaded region under the null distribution curve.

**Both tails count.** A large positive t-value means our estimate is far above the null value; a large negative t-value means it's far below. If we're asking whether the parameter equals some specific value and have no prior reason to prefer one direction over the other, both kinds of deviation are equally informative about whether the null is wrong. This is a two-sided test — the direction doesn't matter, the magnitude does. That's why `lm()` reports `Pr(>|t|)`: the probability that the absolute t-value exceeds what we observed.

**The p-value is a probability under an assumed model.** It answers the question: how compatible are our data with the null model? It does not answer: how likely is the null model to be correct? Those are different questions. A small p-value means our data are not well explained by the null model — not that the null is certainly wrong, and not that the effect is large or practically important. We'll explore those distinctions in Part 8.

## Back to lm()

```{r}
#| label: model
model <- lm(height ~ 1, data = data)
summary(model)
```

`lm()` always tests against 0. For heights, that gives:

$$t = \frac{`r round(sample_mean, 2)` - 0}{`r round(se, 2)`} \approx `r round(t_vs_0, 0)`$$

A t-value of `r round(t_vs_0, 0)` is `r round(t_vs_0, 0)` standard errors from zero. No t-distribution with any reasonable degrees of freedom assigns meaningful probability to values that extreme. The `Pr(>|t|)` column shows `< 2e-16`, which is R's way of saying the number is too small to represent with standard floating-point precision. The three stars (***) are the significance codes at the bottom of the output, indicating a p-value below 0.001.

This result isn't scientifically interesting. Of course the average adult height isn't zero. The t-value of `r round(t_vs_0, 0)` is enormous and obvious. When we add predictors in later posts, the default null hypothesis shifts to "this predictor has no effect", which is a more interesting question.


## Summary

The p-value is the proportion of the null distribution that falls at least as far from zero as our observed t-value. The null distribution is the t-distribution for the null model — the same t-distribution we derived in Part 6, now applied to the assumed parameter rather than our estimate.

For heights tested against 155 cm, p ≈ `r round(p_value_exact, 2)`: we fail to reject the null. Tested against 0 — `lm()`'s default — the p-value is essentially zero, because no one believed the mean height was zero to begin with.

In Part 8, we'll look at what the p-value is not: it's frequently misread as a probability that the null model is correct, a measure of effect size, or a measure of practical importance. None of those are right, and the distinctions matter.
