---
title: "Understanding Regression (Part 7): The p-Value"
description: "We have a t-value and know it follows a t-distribution. In this post, we use that distribution to ask how likely our t-value would be if the null hypothesis were true — and that probability is the p-value."
date: 2026-04-28
categories:
  - statistics
  - tutorial
  - regression
code-fold: true
toc: true
draft: true
---

```{r}
#| label: setup
#| message: false
library(tidyverse)

primary <- "#16a34a"

theme_set(theme_minimal(base_size = 16))
update_geom_defaults(
  "histogram",
  aes(fill = primary, color = "white")
)
```

## Recap

Part 6 introduced the t-value: the distance between our estimate and a reference value, expressed in units of the SE. We also saw that t-values follow the t-distribution rather than the normal, because using an estimated σ adds variability to the SE and produces heavier tails. The last column in the `lm()` output is `Pr(>|t|)`. [this feels disconnected from the previous sentence] This post is about what that number means.

```{r}
#| label: data
data <- read_csv("Howell1.csv")
data <- filter(data, age >= 18)

sample_mean <- mean(data$height)
sample_sd <- sd(data$height)
n <- nrow(data)
se <- sample_sd / sqrt(n)

t_vs_155 <- (sample_mean - 155) / se
t_vs_0 <- sample_mean / se
```

## How likely is our t-value?

We have an estimate and a t-value. The t-value tells us how far our estimate sits from a reference value, in SE units. But knowing the distance doesn't yet tell us whether that distance is large or small in any absolute sense. Is a t-value of 1 surprising? What about 3?

To answer that, we need to know what t-values we'd typically see if the null hypothesis were true — that is, if the reference value actually were the parameter. That distribution is what we can compare our observed t-value against.

Let's start with the following test: does the average height equal 155 cm? Our estimated mean is `r round(sample_mean, 2)` cm, and testing against 155 cm gives:

$$t = \frac{`r round(sample_mean, 2)` - 155}{`r round(se, 2)`} \approx `r round(t_vs_155, 2)`$$

Our estimated mean of `r round(sample_mean, 2)` cm produces a t-value of `r round(t_vs_155, 2)`. Now the question is: if the parameter *were* 155, how often would we expect a t-value at least that far from zero?

[i feel like we should maybe start with a section explaining the logic first, without a specific example]

## The null distribution

We can simulate this directly [we can simulate what directly? I generally don't like continuing from the title as if everybody read the title, at least I don't]. If the parameter were 155, we'd be drawing samples from N(155, σ) [we have not used this notation so far, not consistently at least]. For each simulated sample, we compute t the same way we always do: subtract the reference value and divide by the SE. Repeat this many times to build up the distribution of t-values we'd expect under the null.

```{r}
#| label: fig-null-distribution
#| fig-cap: "Distribution of t-values from 10,000 simulated samples where μ = 155 cm. The dashed curve is the t-distribution; the vertical line marks our observed t-value. The shaded regions show the proportion of t-values at least as extreme as ours."
set.seed(42)

n_simulations <- 10000

null_t_values <- replicate(n_simulations, {
  sim <- rnorm(n, mean = 155, sd = sample_sd)
  (mean(sim) - 155) / (sd(sim) / sqrt(n))
})

x_range <- seq(-5, 5, length.out = 1000)
t_curve <- tibble(x = x_range, y = dt(x_range, df = n - 1))

shade_left <- t_curve |> filter(x <= -abs(t_vs_155))
shade_right <- t_curve |> filter(x >= abs(t_vs_155))

p_value_sim <- mean(abs(null_t_values) >= abs(t_vs_155))
p_value_exact <- 2 * pt(abs(t_vs_155), df = n - 1, lower.tail = FALSE)

ggplot() +
  geom_histogram(
    data = tibble(t = null_t_values),
    aes(x = t, y = after_stat(density)),
    binwidth = 0.1
  ) +
  geom_area(
    data = shade_left,
    aes(x = x, y = y),
    fill = "black",
    alpha = 0.3
  ) +
  geom_area(
    data = shade_right,
    aes(x = x, y = y),
    fill = "black",
    alpha = 0.3
  ) +
  geom_line(
    data = t_curve,
    aes(x = x, y = y),
    linetype = "dashed",
    linewidth = 0.8
  ) +
  geom_vline(
    xintercept = t_vs_155,
    linewidth = 1
  ) +
  labs(x = "t value", y = "Density") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)))
```

The histogram shows the simulated t-values; the dashed curve is the t-distribution with df = `r n - 1`.

The vertical line marks our observed t-value of `r round(t_vs_155, 2)`. Values like that don't look unusual at all; the null distribution produces them regularly. The shaded regions show the proportion of simulated t-values as extreme or more extreme than ours — in both directions, since a t-value of +`r round(abs(t_vs_155), 2)` would be equally surprising as −`r round(abs(t_vs_155), 2)`.

That shaded proportion is the **p-value**. In our simulation, `r round(p_value_sim * 100, 1)`% of t-values fell at least as far from zero as ours did. Using the t-distribution directly [unclear what this means, you mean calculating it from the formula, rather than using simulation] gives p = `r round(p_value_exact, 3)`. Our data are quite compatible with a mean of 155 cm [dangerous phrasing. We should stick close to the meaning of the p-value and signifance logic, we fail to reject the null. We dont want to say this means the null is probably true].

The two-sided framing is why `lm()` reports `Pr(>|t|)`: it's the probability of |t| being at least this large, not just the probability of a t-value this negative [unclear phrasing, I think. 'this negative'].

## Back to lm()

```{r}
#| label: model
model <- lm(height ~ 1, data = data)
summary(model)
```

`lm()` always tests against 0. For heights, that gives:

$$t = \frac{`r round(sample_mean, 2)` - 0}{`r round(se, 2)`} \approx `r round(t_vs_0, 0)`$$

A t-value of `r round(t_vs_0, 0)` is `r round(t_vs_0, 0)` standard errors from zero. No t-distribution with any reasonable degrees of freedom assigns meaningful probability to values that extreme. The `Pr(>|t|)` column shows `< 2e-16`, which is R's way of saying the number is too small to represent with standard floating-point precision. The three stars (***) are the significance codes at the bottom of the output, indicating a p-value below 0.001.

This result isn't scientifically interesting. Of course the average adult height isn't zero. The t-value of `r round(t_vs_0, 0)` is enormous and obvious. When we add predictors in later posts, the default null hypothesis shifts to "this predictor has no effect", which is a more interesting question.

## What the p-value is and isn't

The p-value is frequently misread, so it's worth being precise.

**It is**: the probability of observing a t-value at least as extreme as ours, given that the null hypothesis is true. It measures how compatible our data are with the null.

**It is not** the probability that the null is true. A p-value of `r round(p_value_exact, 2)` from our test against 155 doesn't mean there's a `r round(p_value_exact * 100, 0)`% chance that the mean equals 155. The p-value is a statement about data under an assumed model, not about the model itself.

**It is not** a measure of effect size. With a large enough sample, even a mean of 154.99 cm would produce a very small p-value when tested against 155. The p-value reflects the ratio of the signal to the noise; it doesn't tell you how big the signal is.

**It is not** a measure of practical importance. A small p-value means the data are hard to explain under the null — not that the effect matters in practice.

[perhaps this section is worth its own post, with simulations demonstrating each of these points]

[I guess we will also need a post about treating 5% as significant. this post would be about managing false positives and stuff, about the logic of the decision rule. We don't need to get into this in this post, I think.]

## Summary

The p-value is the probability of seeing a t-value at least as extreme as ours if the null hypothesis were true. We build that null distribution by simulating what t-values would look like under the null, and it turns out to be the t-distribution with n − 1 degrees of freedom — heavier-tailed than the normal, because σ is estimated.

For heights tested against 155 cm, p ≈ `r round(p_value_exact, 2)`: our data are compatible with a mean of 155. Tested against 0 — `lm()`'s default — the p-value is essentially zero, because no one believed the mean height was zero to begin with.

In Part 8, we'll turn to the third goal from Part 1: prediction.
