---
title: "Figuring out Bayesian statistics (Part 1)""
description: "Bayesian statistics seems pretty cool, but I don't really know how to apply it yet. In this blog post, I try to setup a Bayesian workflow that teaches both you and me how to do it."
date: 2022-11-24
categories:
  - statistics
  - tutorial
  - Bayesian statistics
  - regression
code-fold: true
code-tools: true
toc: true
format: 
   html:
     df-print: kable
draft: true
---

```{r}
#| label: setup
#| message: false

# Load packages
library(tidyverse)
library(viridis)
library(brms)
library(tidybayes)

# Load data
data <- read_csv("Howell1.csv")

# Set seed
set.seed(4)

# Set options
theme_set(theme_minimal())
```

This post is about developing a workflow that I think makes sense for analyzing data using Bayesian statistics.

The data we will use to play with is the same data Richard McElreath uses in Chapter 4 of his amazing book called [Statistical Rethinking](https://xcelab.net/rm/statistical-rethinking/ "Statistical Rethinking website"). The data consists of partial census data of the !Kung San, compiled from interviews conducted by Nancy Howell in the late 1960s. Just like in the book, we will focus only on people 18 years or older, so in the code below we create a subset of the data and store the result in a data frame called `data`.

```{r}
#| label: data
#| tbl-cap: Partial census data for the Dobe area !Kung San compiled by Nancy Howell in the late 1960s.

# Only select people older than 18
data <- filter(data, age >= 18)

# Show the first rows
head(data)
```

The general idea behind Bayesian statistics is that you start with some prior beliefs about the parameters of interest and then update those beliefs with the data. Note that this doesn't mean that you have to personally accept those beliefs. You could simply postulate a belief to serve a particular purpose, such as assuming that a null effect is most likely even though you personally believe that there should be an effect. It does mean that when we want to analyze the data, we should start with defining our beliefs, rather than immediately jumping into running an analysis.

Let's focus our first question on the heights in the data. We should begin by defining a belief that describes the different heights, which is based on our a priori knowledge of the heights of the Dobe area !Kung San. In other words, we have to describe what we believe their heights to be. This is unlike what you have to do with frequentist statistics, so this part might be a bit tricky.

To make it easier, we will use the amazing `brms` package to both define and inspect our beliefs, as well as use the data to update those beliefs.

## An intercept-only model

If we start with analyzing only the heights data then we'll be constructing an intercept-only model. You may be familiar with the R formula for this type of model: `height ~ 1`.

With this formula and the data we can use `brms` to figure out which priors we need to set by running the `get_prior()` function. This is probably the easiest way to figure which priors you need when you're just starting out using `brms`.

```{r}
#| label: get-prior
get_prior(height ~ 1, data = data)
```

The output shows us that we need to set two priors, one for the Intercept and one for sigma.

However, this is not the best way to think about the model. Using the function will give you the answer, but it doesn't really improve our understanding of why we need these two priors. In this case we also omitted an important specification of the heights, which is that we think they are normally distributed (the default). So let's instead write down our model in a different way, which explicitly specifies how we think the heights are distributed and which parameters we need to set priors on. If we think the heights are normally distributed, we define our model like this:

> $heights_i ∼ Normal(\mu, \sigma)$

We explicitly note that the heights come from a normal distribution, which is determined by two parameters: $\mu$ and $\sigma$. This then also immediately tells us that we need to set two priors, one on $\mu$ and one on $\sigma$.

In our intercept-only model, the $\mu$ parameter refers to our intercept and the $\sigma$ parameter refers to, well, sigma. It's not often discussed in the literature I'm familiar with, but we'll figure it out below. In fact, let's discuss each of these parameters in turn and figure out what kind of prior makes sense.

## The Intercept prior ($\mu$)

The prior for the intercept indicates what we believe the *average* height of the !Kung San to be.

`brms` has set the default Intercept prior as a Student *t*-distribution with 3 degrees of freedom, a mean of 154.3 and a standard deviation of 8.5. That means `brms` starts off with a 'belief' that the *average* of the heights is roughly normally distributed, with the most common average height being 154.3, but with quite some uncertainty. In fact, a Student *t*-distribution has thicker tails compared to a normal distribution, which means more uncertainty, at least at low sample sizes. With these values you think the average height is most likely to be 154.3, but you also think that quite some other values are possible, even much smaller or much taller average heights.

But this is the default prior. `brms` determines this automatic prior by peeking at the data, which is not what we want to do. Instead, we should create our own.

So what do I believe the average height to be? As a Dutch person, I might be under the impression that the average height is around 175 centimeters. This is probably too tall to use as an average for the !Kung San because we're known for being quite tall. So I think the average should be lower than 175, perhaps 170. I am not very sure, though. After all, I am far from an expert on people's heights; I am only using my layman knowledge here. As a result, an average of 165 seems possible to me too. So let's describe my belief in the form of a distribution in which multiple averages are possible, to varying extents. We could use different types of distributions for this purpose. We could use a Student *t*-distribution, but we can also use a normal distribution. We should use a Student *t*-distribution with small degrees of freedom if we want to allow for the possibility of being very wrong (remember, it has thicker tails, so it covers a wider range of average heights). We're not super uncertain about people's heights, though, so let's use a normal distribution.

As we saw in defining our height model, a normal distribution requires that we set the $\mu$ and the $\sigma$. The $\mu$ we already covered (i.e., 170), so that leaves $\sigma$. Let's set this to 10 and see what happens by visualizing this prior. Below I plot both the default `brms` prior and our own.

```{r}
#| label: height-mu-prior
#| fig-cap: Two priors for $\mu$\: brms' default and my own
height_prior_intercept <- tibble(
  height_mean = seq(from = 100, to = 250, by = 0.1),
  ours = dnorm(height_mean, mean = 170, sd = 10),
  default = dstudent_t(height_mean, df = 30, mu = 154.3, sigma = 8.5),
)

height_prior_intercept <- pivot_longer(
  height_prior_intercept, 
  cols = -height_mean, 
  names_to = "prior"
) 

ggplot(
    height_prior_intercept, 
    aes(x = height_mean, y = value, linetype = fct_rev(prior))
  ) +
  geom_line() +
  labs(x = "Average height", y = "", linetype = "Prior") +
  scale_x_continuous(breaks = seq(100, 250, 20))
```

Our prior indicates that we believe the average height to be higher than the default prior. In terms of the standard deviation, we both seem to be about equally uncertain about this average. To be fair, I think this prior of ours is not very plausible. Apparently we assign quite a chunk of plausibility to an average of 180 cm, or even 190 cm, which is very unlikely. An average of 160 cm is more plausible to me than an average of 180, so I should probably lower the mu, or use more of a skewed distribution. This is one of the benefits of visualizing the prior, it lets you think again about your prior so that you may improve on it. Regardless, we can keep the prior like this for now. We'll see later that our data easily overshadows our prior.

### The sigma prior

What about the standard deviation? I find setting the standard deviation of the distribution of heights (not the mean of the heights) quite difficult. There are parts that are easy, such as the fact that the standard deviation has to be 0 or larger (it can't be negative), but exactly how large it should be, I don't know.

I do know it is unlikely to be close to 0, and unlikely to be very large. That's because I know people's heights do vary, so I know the sigma can't be 0. I also know it's not super large because we don't see people who are taller than 2 meters very often. This means the peak should be somewhere above 0, with a tail to allow higher values but not too high. We can use a normal distribution for this with a mean above 0 and a particular standard deviation, and ignore everything that's smaller than 0.

As I mentioned before, there is a downside of using a normal distribution, though. Normal distributions have long tails, but there is actually very little density in those tails. If we are quite uncertain about our belief about sigma, we should have thicker tails to indicate we believe those values are more plausible. One way to do this is by using a Student *t*-distribution, perhaps even a cauchy distribution. Actually, the cauchy distribution is a special case of the Student *t* distribution; they are equivalent if the degree of freedom is 1. By increasing the degrees of freedom, you make the *t*-distribution more similar to a normal distribution.

So, a *t*-distribution requires three parameters: $\mu$, $\sigma$, and the degrees of freedom. I set $\mu$ to 5,$\sigma$ to 5, and the degrees of freedom to 1. Below I plot this prior and `brms`'s default prior.

```{r}
#| label: height-sigma-prior
#| fig-cap: Two priors for $\sigma$ - brms' default and my own
height_prior_sigma <- tibble(
  height_sigma = seq(from = 0, to = 50, by = .1),
  default = dstudent_t(height_sigma, df = 3, mu = 0, sigma = 8.5) * 2,
  ours = dstudent_t(height_sigma, df = 1, mu = 5, sigma = 5) * 2
)

height_prior_sigma <- pivot_longer(
  height_prior_sigma, 
  cols = -height_sigma, 
  names_to = "prior"
)

ggplot(
    height_prior_sigma, 
    aes(x = height_sigma, y = value, linetype = fct_rev(prior))
  ) +
  geom_line() +
  labs(x = "Standard deviation of heights", y = "", linetype = "Prior")
```

As you can see, both distributions have long tails, allowing for the possibility of high standard deviations. There are some notable differences between the two priors. Our prior puts more weight on a standard deviation larger than 0, while the default prior reflects a belief in which a standard deviation of 0 is most likely. However, both priors are quite weak. We'll see that the data easily overshadows these priors.

Before we run the analysis, we can also check the results of both our priors on the distribution of heights.

### A prior predictive check

Before we run our model, we should check what the effect is of both priors combined. Because we have set the priors we can simulate what we believe the data to be. This is one way to see whether our priors actually make sense. It is called a prior predictive check.

We can use `brms` to do this by running the `brm()` function. However, instead of running the actual model, we tell it to only sample from the prior.

```{r}
#| label: height-prior
model_height_prior <- brm(
  height ~ 1,  
  data = data, 
  family = gaussian,
  prior = c(
      prior(normal(170, 10), class = "Intercept"),
      prior(cauchy(5, 5), class = "sigma")
    ), 
  cores = 4,
  seed = 4, 
  sample_prior = "only",
  file = "models/model_height_prior.rds"
)

data <- mutate(data, height_z = (height - mean(height)) / sd(height))

model_height_prior <- brm(
  height ~ 1,  
  data = data, 
  family = gaussian,
  prior = c(
      prior(normal(170, 10), class = "Intercept"),
      prior(cauchy(5, 5), class = "sigma")
    ), 
  cores = 4,
  seed = 4, 
  sample_prior = "only",
  file = "models/model_height_prior.rds"
)
```

We then use the `tidybayes` package to draw samples from the prior and plot these draws.

```{r}
#| label: prior-predictive
#| fig-cap: Prior predictive check
predictions_prior <- tibble(distribution = "prior")

predictions_prior <- add_predicted_draws(
  newdata = predictions_prior, 
  object = model_height_prior, 
  value = "predicted_height"
)

ggplot(predictions_prior, aes(x = predicted_height)) +
  geom_histogram(binwidth = 1, alpha = .85) +
  xlim(100, 250) +
  labs(x = "Height", y = "")
```

So, our priors result in a normal distribution of heights ranging from about 125 cm to 225 cm. That is probably a bit too wide, but let's go with this. Let's run the model for real now.

```{r}
#| label: intercept-model
model_height <- brm(data = data, 
  family = gaussian,
  height ~ 1,
  prior = c(
    prior(normal(170, 10), class = "Intercept"),
    prior(cauchy(5, 5), class = "sigma")
  ),
  cores = 4,
  seed = 4,
  sample_prior = TRUE,
  file = "models/model_height.rds"
)
```

Before we check the results, let's first check whether the chains look good.

```{r}
#| label: chains
plot(model_height)
```

It seems like they do. The distributions look normal and the chains look like caterpillars, which means they're sampling from the distribution space and that's what we want. This also shows us the posterior distributions of the two parameters, but let's create our own graphs that compare both the prior and posterior distributions.

```{r}
#| label: prior-posterior-mu
#| fig-cap-: Prior vs. posterior for $\mu$
results <- model_height %>%
  gather_draws(b_Intercept, sigma, prior_Intercept, prior_sigma) %>%
  ungroup() %>%
  mutate(
    parameter = if_else(str_detect(.variable, "sigma"), "sigma", "intercept"),
    distribution = if_else(str_detect(.variable, "prior"), "prior", "posterior")
  )

results_intercept <- filter(results, parameter == "intercept")
results_sigma <- filter(results, parameter == "sigma")

ggplot(results_intercept, aes(x = .value, fill = fct_rev(distribution))) +
  geom_histogram(binwidth = 1, position = "identity", alpha = .85) +
  xlim(145, 195) +
  labs(x = "Average height", y = "", fill = "Distribution") +
  scale_fill_viridis(option = "mako", discrete = TRUE, begin = .25, end = .75)
```

Here we see that the posterior distribution of average heights is now much more narrow and centered around 156 cm.

```{r}
#| label: prior-posterior-sigma
#| fig-cap-: Prior vs. posterior for $\sigma$
ggplot(results_sigma, aes(x = .value, fill = fct_rev(distribution))) +
  geom_histogram(binwidth = 0.25, position = "identity", alpha = .85) + 
  xlim(0, 25) +
  labs(x = "Height standard deviation", y = "", fill = "Distribution") +
  scale_fill_viridis(option = "mako", discrete = TRUE, begin = .25, end = .75)
```

Similarly, we see that the posterior for sigma is also much more narrow and around 7.75.

We can call up the estimates and the 95% confidence intervals by printing the model object.

```{r}
#| label: model-height
model_height
```

Our final step is to visualize the posterior distribution of all heights and compare it the distribution of heights based on our priors (the prior predictive check).

```{r}
#| label: prior-posterior-predictive-check
#| fig-cap: Prior and posterior predictive check
predictions_posterior <- tibble(distribution = "posterior")

predictions_posterior <- add_predicted_draws(
  newdata = predictions_posterior,
  object = model_height, 
  value = "predicted_height"
)

predictions <- bind_rows(predictions_prior, predictions_posterior)

ggplot(predictions, aes(x = predicted_height, fill = distribution)) +
  geom_histogram(binwidth = 1, alpha = .85) +
  xlim(100, 250) +
  labs(x = "Height", y = "", fill = "Distribution") +
  scale_fill_viridis(option = "mako", discrete = TRUE, begin = .25, end = .75)
```

Based on the data, we should now believe, after seeing the data, that people's heights are lower than expected.

## Adding a predictor

Let's repeat the same steps but this time we add a predictor to the model. We'll add weight as a predictor to see its relationship with height.

The formula becomes: `height ~ 1 + weight`.

So let's see which priors we need to set for this model.

```{r}
#| label: height-weight-default-priors
get_prior(height ~ 1 + weight, data = data)
```

This time the output is a bit more confusing. We seem to have two additional priors: a default b prior and a vectorized b prior for the weight coefficient. The reason there are now two additional priors is that there are two different ways to specify a prior here. We can either set a specific prior for the weight coefficient, or set a prior on all the b-class priors. In this specific case, with only one additional predictor, it means we can use either technique and we only need to set one of them.

What should our prior for weight be? In other words, for every increase of 1 in weight, how much do we think this relates to an increase (or decrease) in height? This is likely to be a positive relationship (taller people are more likely to be heavier as well), but we can also refrain from indicating the direction. This means we can specify a distribution of possible effects, centered around 0. The only decision we're left with is setting the standard deviation. Let's try a value of 10 and see what this means.

```{r}
#| label: weight-prior
model_height_weight_prior <- brm(
  height ~ 1 + weight,  
  data = data, 
  family = gaussian,
  prior = c(
      prior(normal(170, 10), class = "Intercept"),
      prior(cauchy(5, 1), class = "sigma"),
      prior(normal(0, 10), coef = "weight")
    ),
  cores = 4,
  seed = 4, 
  sample_prior = "only",
  file = "models/model_height_weight_prior.rds"
)
```

We can do our prior predictive checking in two different ways now. We can check only the effect of our prior on weight, or also include the prior on sigma.

In the former case, we use `fitted_draws()` to obtain draws from our posterior, like so:

```{r}
draws <- add_predicted_draws(
  tibble(weight = seq(30, 100, 1)),   
  model_height_weight_prior,
  value = "fitted_height",
  ndraws = 100
) 

ggplot(draws, aes(x = weight, y = fitted_height, group = .draw)) +
  geom_line(alpha = .5) 
```

```{r}
samples <- as_draws_df(model_height_weight_prior)

ggplot(samples, aes(x = b_weight)) +
  geom_histogram()
```

In both cases, we see that we get quite some implausible values. So let's take a prior that produces fewer crazy observations, such as a prior of `normal(0, 1)`.

```{r height_weight_prior2}
model_height_weight_prior2 <- brm(
  height ~ 1 + weight,  
  data = data, 
  family = gaussian,
  prior = c(
      prior(normal(170, 10), class = "Intercept"),
      prior(cauchy(5, 1), class = "sigma"),
      prior(normal(0, 1), coef = "weight")
    ),
  cores = 4,
  seed = 4, 
  sample_prior = "only",
  file = "models/model_height_weight_prior2.rds"
)

model_height_weight_prior2 %>%
  fitted_draws(tibble(weight = seq(30, 100, 1)), value = "fitted_height", n = 100) %>%
  ggplot(aes(x = weight, y = fitted_height, color = .draw, group = .draw)) +
    geom_line(alpha = .5)
```

That looks much better. It could still be refined because it produces several highly implausible results, but let's stick with this for now and run the model.

```{r height_weight, echo = TRUE}
model_height_weight <- brm(
    height ~ 1 + weight,  
    data = data, 
    family = gaussian,
    prior = c(
        prior(normal(170, 10), class = "Intercept"),
        prior(cauchy(5, 1), class = "sigma"),
        prior(normal(0, 1), coef = "weight")
      ),
    cores = 4,
    seed = 4, 
    sample_prior = TRUE,
    file = "models/model_height_weight.rds"
  )
```

Let's check the chains.

```{r}
plot(model_height_weight)
```

And let's compare the prior and posterior distribution of our weight predictor.

```{r}
results <- model_height_weight %>%
  posterior_samples() %>%
  select(prior_b_weight, b_weight) %>%
  pivot_longer(cols = everything()) %>%
  mutate(
    distribution = if_else(str_detect(name, "prior"), "prior", "posterior")
  ) %>%
  select(-name)

ggplot(results, aes(x = value, fill = distribution)) +
  geom_histogram(binwidth = 0.1) +
  coord_cartesian(xlim = c(-3, 3)) +
  labs(x = "Average height", y = "", fill = "Distribution")
```

Again our prior was quite broad and our posterior is much more narrow. In fact, the estimates are:

```{r}
model_height_weight
```

Weight is positively related with height (0.90), with a 95% CI ranging from 0.82 to 0.99.

## Summary

The workflow consists of the following steps:

1.  Specify your model
2.  Check which priors are necessary by looking at the parameters in your model
3.  Set priors
4.  Run `brms()` and sample from the prior only
5.  Perform prior predictive checks to see whether the priors produce plausible values
6.  Adapt the priors if needed
7.  Run the model with the final priors
8.  Check the chains
9.  Inspect the model output
10. Optional: Compare priors to posteriors to see how your beliefs have updated
