---
title: "Figuring out Bayesian statistics"
description: "Bayesian statistics seems pretty cool, but I don't really know how to apply it yet. In this blog post, I try to setup a Bayesian workflow that teaches both you and me how to do it."
date: 2022-11-24
categories:
  - statistics
  - tutorial
  - Bayesian statistics
  - regression
code-fold: true
code-tools: true
toc: true
format: 
   html:
     df-print: kable
---

::: callout-warning
This post is a work in progress. I already put it up to receive some feedback, so expect changes to the content of this post.
:::

This post is about figuring out how Bayesian statistics works and about developing a workflow to conduct Bayesian analyses using the `brms` package. Specifically, I want to go through the process of running a Bayesian analysis and visualizing the different steps of that analysis in order to make sure I know what I'm doing. With a bit of luck this is educational and we'll both end up learning how this works!

If you want to follow a long, run the following setup code.

```{r}
#| label: setup
#| message: false

# Load packages
library(tidyverse)
library(viridis)
library(brms)
library(tidybayes)

# Load data
data <- read_csv("Howell1.csv")

# Set seed
set.seed(4)

# Set options
theme_set(theme_minimal())
```

The data we will use to play with is the same data Richard McElreath uses in Chapter 4 of his amazing book called [Statistical Rethinking](https://xcelab.net/rm/statistical-rethinking/ "Statistical Rethinking website"). The data consists of partial census data of the !Kung San, compiled from interviews conducted by Nancy Howell in the late 1960s. Just like in the book, we will focus only on people 18 years or older, so in the code below we create a subset of the data and store the result in a data frame called `data`.

```{r}
#| label: data
#| tbl-cap: Partial census data for the Dobe area !Kung San compiled by Nancy Howell in the late 1960s.

# Only select people older than 18
data <- filter(data, age >= 18)

# Show the first rows
head(data)
```

The general idea behind Bayesian statistics is that you start with some prior beliefs about the parameters of interest and then update those beliefs with the data. Note that this doesn't mean that you have to personally accept those beliefs. You could simply postulate a belief to serve a particular purpose, such as assuming that a null effect is most likely even though you personally believe that there should be an effect. It does mean that we should start with defining some beliefs, rather than immediately jumping into running an analysis.

Let's focus our first question on the heights in the data. What are the heights of the Dobe area !Kung San?

## An intercept-only model

If we start with analyzing only the heights data then we'll be constructing an intercept-only model. You may be familiar with the R formula for this type of model: `height ~ 1`.

With this formula and the data we can use `brms` to figure out which priors we need to set by running the `get_prior()` function. This is probably the easiest way to figure which priors you need when you're just starting out using `brms`.

```{r}
#| label: get-prior
get_prior(height ~ 1, data = data)
```

The output shows us that we need to set two priors, one for the Intercept and one for sigma. `brms` already determined a default prior for each, but we'll want to construct our own. `brms` offers these default priors so you can immediately run an analysis if you want to (they are required for a Bayesian analysis), but it is recommended to construct your own.

Using the `get_prior()` function is not the best way to think about which priors we need, though. Using the function will give you the answer, but it doesn't really improve our understanding of why we need these two priors. In this case we also omitted an important specification of the heights, which is that we think they are normally distributed (the default assumption in `get_prior()`). So let's instead write down our model in a different way, which explicitly specifies how we think the heights are distributed and which parameters we need to set priors on. If we think the heights are normally distributed, we define our model like this:

$$heights_i ∼ Normal(\mu, \sigma)$$

We explicitly note that the heights come from a normal distribution, which is determined by the parameters $\mu$ and $\sigma$. This then also immediately tells us that we need to set two priors, one on $\mu$ and one on $\sigma$.

In our intercept-only model, the $\mu$ parameter refers to our intercept and the $\sigma$ parameter refers to sigma. sigma is not often discussed in the literature I'm familiar with, but we'll figure it out below. In fact, let's discuss each of these parameters in turn and figure out what kind of prior makes sense.

### The Intercept ($\mu$) prior

The prior for the intercept indicates what we believe the *average* height of the !Kung San to be.

`brms` has set the default Intercept prior as a Student *t*-distribution with 3 degrees of freedom, a mean of 154.3 and a standard deviation of 8.5. That means `brms` starts off with a 'belief' that the *average* of the heights is 154.3, but with quite some uncertainty reflected in the standard deviation of 8.5 and the fact that the distribution is a Student *t*-distribution. A Student *t*-distribution has thicker tails compared to a normal distribution, meaning that numbers in the tails of the distribution are considered more likely compared to a normal distribution, at least when the degrees of freedom are low. At higher degrees of freedom, the *t*-distribution becomes more and more like the normal distribution. So, the thicker tails of the *t*-distributions means smaller and taller average heights are relatively more plausible.

But this is the default prior. `brms` determines this automatic prior by peeking at the data, which is not what we want to do. Instead, we should create our own.

So what do I believe the average height to be? As a Dutch person, I might be under the impression that the average height is around 175 centimeters. This is probably too tall to use as an average for the !Kung San because we're known for being quite tall. So I think the average should be lower than 175, perhaps 170. I am not very sure, though. After all, I am far from an expert on people's heights; I am only using my layman knowledge here. An average of 165 seems possible to me too. So let's describe my belief in the form of a distribution in which multiple averages are possible, to varying extents. We should use a Student *t*-distribution with small degrees of freedom if we want to allow for the possibility of being very wrong (remember, it has thicker tails, so it assigns more probability to a wider range of average heights). We're not super uncertain about people's heights, though, so let's use a normal distribution.

As we saw in defining our height model, a normal distribution requires that we set the $\mu$ and the $\sigma$. The $\mu$ we already covered (i.e., 170), so that leaves $\sigma$. Let's set this to 10 and see what happens by visualizing this prior. Below I plot both the default `brms` prior and our own with $\mu$ = 170 and $\sigma$ = 10.

```{r}
#| label: height-mu-prior
#| fig-cap: Two priors for $\mu$
height_prior_intercept <- tibble(
  height_mean = seq(from = 100, to = 250, by = 0.1),
  ours = dnorm(height_mean, mean = 170, sd = 10),
  default = dstudent_t(height_mean, df = 30, mu = 154.3, sigma = 8.5),
)

height_prior_intercept <- pivot_longer(
  height_prior_intercept, 
  cols = -height_mean, 
  names_to = "prior"
) 

ggplot(
    height_prior_intercept, 
    aes(x = height_mean, y = value, linetype = fct_rev(prior))
  ) +
  geom_line() +
  labs(x = "Average height", y = "", linetype = "Prior") +
  scale_x_continuous(breaks = seq(100, 250, 20))
```

Our prior indicates that we believe the average height to be higher than the default prior. In terms of the standard deviation, we both seem to be about equally uncertain about this average. To be fair, I think this prior of ours is not very plausible. Apparently we assign quite a chunk of plausibility to an average of 180 cm, or even 190 cm, which is very unlikely. An average of 160 cm is more plausible to me than an average of 180, so I should probably lower the mu, or use more of a skewed distribution. This is one of the benefits of visualizing the prior, it lets you think again about your prior so that you may improve on it. Regardless, we can keep the prior like this for now. We'll see later that our data easily overshadows our prior.

### The sigma ($\sigma$) prior

What about the sigma prior? What even is sigma? Sigma is the estimated standard deviation of the errors or standard deviation of the residuals of the model. In the simple case of an intercept-only model, this is identical to what we believe the standard deviation of the heights to be.

I find setting the standard deviation of the distribution of heights (not the mean of the heights) quite difficult. There are parts that are easy, such as the fact that the standard deviation has to be 0 or larger (it can't be negative), but exactly how large it should be, I don't know.

I do know it is unlikely to be close to 0, and unlikely to be very large. That's because I know people's heights do vary, so I know the sigma can't be 0. I also know it's not super large because we don't see people who are taller than 2 meters very often. This means the peak of our prior should be somewhere above 0, with a tail to allow higher values but not too high. We can use a normal distribution for this with a mean above 0 and a particular standard deviation, and ignore everything that's smaller than 0 (`brms` automatically ignores negative values for $\sigma$).

As I mentioned before, there is a downside of using a normal distribution, though. Normal distributions have long tails, but there is actually very little density in those tails. If we are quite uncertain about our belief about sigma, we should use a *t*-distribution, or perhaps even a cauchy distribution (actually, the cauchy distribution is a special case of the Student *t*-distribution; they are equivalent if the degree of freedom is 1). The lower the degrees of freedom, the more probability we assign to higher and lower values.

So, a *t*-distribution requires three parameters: $\mu$, $\sigma$, and the degrees of freedom. I set $\mu$ to 5, $\sigma$ to 5, and the degrees of freedom to 1. Below I plot this prior and `brms`'s default prior to get a better grasp of these priors.

```{r}
#| label: height-sigma-prior
#| fig-cap: Two priors for $\sigma$
height_prior_sigma <- tibble(
  height_sigma = seq(from = 0, to = 50, by = .1),
  default = dstudent_t(height_sigma, df = 3, mu = 0, sigma = 8.5),
  ours = dstudent_t(height_sigma, df = 1, mu = 5, sigma = 5) 
)

height_prior_sigma <- pivot_longer(
  height_prior_sigma, 
  cols = -height_sigma, 
  names_to = "prior"
)

ggplot(
    height_prior_sigma, 
    aes(x = height_sigma, y = value, linetype = fct_rev(prior))
  ) +
  geom_line() +
  labs(x = "Standard deviation of heights", y = "", linetype = "Prior")
```

As you can see, both distributions have longish tails, allowing for the possibility of high standard deviations. There are some notable differences between the two priors, though. Our prior puts more weight on a standard deviation larger than 0, while the default prior reflects a belief in which a standard deviation of 0 is most likely. However, both priors are quite weak. We'll see that the data easily overshadows these priors.

So far we have inspected each prior in isolation, but we can also inspect the results of our priors combined. This is called a prior predictive check.

### Prior predictive check

With our priors we can simulate what we believe the data to be. This is one way to see whether our priors actually make sense. It is called a prior predictive check.

We can use `brms` to do this by running the `brm()` function. The `brm()` function is the main work horse of the `brms` package. It allows us to run Bayesian analyses by using a notation style that you may be familiar with if you experience with running analyses in R. This is also one of the reasons why the `brms` package is so great; it's so easy to get started with running Bayesian analyses.

The `brm()` function requires a model specification (using a formula, just like in `lm()`) and the data. Optionally, but usefully, we should also specify the response distribution (a normal distribution by default) and the priors (remember, we should not rely on the default priors).

However, we're not ready to actually run the model just yet. Instead, we will kinda trick `brms` into running an analysis, but tell it to only sample from the prior using the `sample_prior` argument. This will give us 'predicted' responses based entirely on our priors and not the data.

```{r}
#| label: height-prior
model_height_prior <- brm(
  height ~ 1,  
  data = data, 
  family = gaussian,
  prior = c(
    prior(normal(170, 10), class = "Intercept"),
    prior(cauchy(5, 5), class = "sigma")
  ), 
  cores = 4,
  seed = 4, 
  sample_prior = "only",
  file = "models/model_height_prior.rds"
)
```

We then use the `tidybayes` package to draw samples from the prior and plot these draws.

```{r}
#| label: prior-predictive
#| fig-cap: Prior predictive check
#| warning: false
predictions_prior <- tibble(distribution = "prior") %>%
  add_predicted_draws(
    object = model_height_prior, 
    value = "predicted_height"
  )

ggplot(predictions_prior, aes(x = predicted_height)) +
  geom_histogram(binwidth = 1, alpha = .85) +
  xlim(100, 250) +
  labs(x = "Height", y = "")
```

So, our priors result in a normal distribution of heights ranging from about 125 cm to 225 cm. That is too wide, but let's run the model to see what happens.

### Running the model

We run the model with the code below. Notice that we still sample from the prior so we can not only visualize our posterior later, but also the priors we have just defined.

```{r}
#| label: intercept-model
model_height <- brm(data = data, 
  family = gaussian,
  height ~ 1,
  prior = c(
    prior(normal(170, 10), class = "Intercept"),
    prior(cauchy(5, 5), class = "sigma")
  ),
  cores = 4,
  seed = 4,
  sample_prior = TRUE,
  file = "models/model_height.rds"
)
```

After running the model, we first check whether the chains look good. We want the chains to look like caterpillars because that indicates we have samples from the entire distribution space of the posterior (and prior).

```{r}
#| label: chains
plot(model_height)
```

The chains look good.

We can call up the estimates and the 95% confidence intervals by printing the model object.

```{r}
#| label: model
summary(model_height)
```

Here we see the Intercept and sigma estimates. Apparently our posterior estimate for the Intercept is `r round(mean(as_tibble(model_height)$b_Intercept), 2)` and the estimate for $\sigma$ is `r round(mean(as_tibble(model_height)$sigma), 2)`. We also see the 95% CIs, but let's visualize these results instead.

Inspecting the chains also showed us the posterior distributions of the two parameters, but let's create our own graphs that compare both the prior and posterior distributions.

```{r}
#| label: prior-posterior-mu
#| fig-cap-: Prior vs. posterior for $\mu$
#| warning: false
results <- model_height %>%
  gather_draws(b_Intercept, sigma, prior_Intercept, prior_sigma) %>%
  mutate(
    parameter = if_else(str_detect(.variable, "sigma"), "sigma", "intercept"),
    distribution = if_else(str_detect(.variable, "prior"), "prior", "posterior")
  )

results_intercept <- filter(results, parameter == "intercept")
results_sigma <- filter(results, parameter == "sigma")

ggplot(results_intercept, aes(x = .value, fill = fct_rev(distribution))) +
  geom_histogram(binwidth = 1, position = "identity", alpha = .85) +
  xlim(145, 195) +
  labs(x = "Average height", y = "", fill = "Distribution") +
  scale_fill_viridis(option = "mako", discrete = TRUE, begin = .25, end = .75)
```

Here we see that the posterior distribution of average heights is now much more narrow and centered around `r round(mean(as_tibble(model_height)$b_Intercept))` cm. So not only should we switch from thinking the average is a lot lower than 170, we can also be much more confident about the mean.

How about sigma?

```{r}
#| label: prior-posterior-sigma
#| fig-cap-: Prior vs. posterior for $\sigma$
#| warning: false
ggplot(results_sigma, aes(x = .value, fill = fct_rev(distribution))) +
  geom_histogram(binwidth = 0.25, position = "identity", alpha = .85) + 
  xlim(0, 25) +
  labs(x = "Height standard deviation", y = "", fill = "Distribution") +
  scale_fill_viridis(option = "mako", discrete = TRUE, begin = .25, end = .75)
```

Similarly, we see that the posterior for sigma is also much more narrow and around `r round(mean(as_tibble(model_height)$sigma))`.

A final step is to visualize the posterior distribution of all heights (a posterior predictive check) and compare it the distribution of heights based on our priors (the prior predictive check).

```{r}
#| label: prior-posterior-predictive-check
#| fig-cap: Prior and posterior predictive check
#| warning: false
predictions_posterior <- tibble(distribution = "posterior")

predictions_posterior <- add_predicted_draws(
  newdata = predictions_posterior,
  object = model_height, 
  value = "predicted_height"
)

predictions <- bind_rows(predictions_prior, predictions_posterior)

ggplot(predictions, aes(x = predicted_height, fill = distribution)) +
  geom_histogram(binwidth = 1, alpha = .85, position = "identity") +
  xlim(100, 250) +
  labs(x = "Height", y = "", fill = "Distribution") +
  scale_fill_viridis(option = "mako", discrete = TRUE, begin = .25, end = .75)
```

## Adding a predictor

Now let's add a predictor to our model. Besides heights, the data set also contains weights. We can create a model in which we regress heights onto weights. The formula syntax for a model like that in R is `height ~ weight`. We can use this formula again in `get_prior()` to see which priors we need to specify.

```{r}
#| label: get-priors-predictor
get_prior(height ~ weight, data = data)
```

The output is a bit trickier this time. We see the Intercept and sigma priors from our previous model, as well as two extra rows referring to a class called `b`. These two rows actually refer to the same prior, one refers specifically to the weight predictor and one refers to all predictors. If you run a model with many more predictors, you could set one prior that applies to all predictors. In this case though, we only have 1 predictor so it actually doesn't matter, both refer to the same prior.

Given that this is a bit trickier, and given that I said writing down your model explicitly is better, we should go ahead and do that.

$$heights_i ∼ Normal(\mu_i, \sigma)$$

$$
\mu_i = \alpha + \beta x_i
$$

We again specify that the heights are normally distributed, so we still have a $\mu$ and $\sigma$, but this time the $\mu$ is no longer a parameter we will estimate. Instead, it's constructed from other parameters, $\alpha$, $\beta$, and an observed variable $x_i$ (the weight observations).

If you're used to linear regression equations, this notation should not surprise you. $\alpha$ refers to the intercept and $\beta$ to the slope.

We need to set priors on these parameters. The prior for $\alpha$ can be the same as the prior for $\mu$ from the previous intercept-only mode, if we center the data so the intercept refers to the average height of someone with an average weight, rather than someone with 0 weight (the default, which makes no sense). So let's first mean center the weight observations.

```{r}
#| label: mean-center-weight
data <- mutate(data, weight_mc = weight - mean(weight))
```

Now we can use the same prior as before, which was a normal distribution with a mean of 170 and a standard deviation of 10 (assuming we did not update this as a result of the previous analysis).

Next is the prior for the slope. This represents the relationship between weights and heights. For every 1 increase in weight, how much do we think that the height will increase or decrease? We could begin with an agnostic prior in which we do not specify the direction and instead just add some uncertainty so the slope can go in either direction. For example, let's put a normal distribution on the slope with a mean of 0 and a standard deviation of 10.

Finally, we have the prior for sigma ($\sigma$). To remind you, sigma refers to the standard deviation of the errors or the residual standard deviation. Now that we have a predictor that means the sigma can be less than what it was in the intercept-only model because some of the variance in heights might be explained by the weights, thus decreasing the size of the residuals and reducing sigma. So, if we believe in a relationship between heights and weights, we should change our prior for sigma so that it's lower. Given that we used a prior for the slope that is agnostic (there could be a positive, negative, or no relationship), our prior for sigma could be left unchanged because it was broad enough to allow for these possibilities.

### Prior predictive check

We can again create a prior predictive check to see whether our priors actually make sense. However, instead of plotting the predicted distribution of heights, we're mostly interested in the relationship between weight and height, so we should plot a check of that relationship instead. We could simulate our own data like we did in the previous section or we can just run the Bayesian model and only draw from the prior, which we also did in the previous section.

```{r}
#| label: weight-prior
#| fig-cap: Several priors for $\beta$
model_height_weight_prior <- brm(
  height ~ weight_mc,  
  data = data, 
  family = gaussian,
  prior = c(
      prior(normal(170, 10), class = "Intercept"),
      prior(cauchy(5, 5), class = "sigma"),
      prior(normal(0, 10), class = "b")
    ), 
  cores = 4,
  seed = 4, 
  sample_prior = "only",
  file = "models/model_height_weight_prior.rds"
)
```

Below we draw intercepts and slopes from the model result and plot 100 of them. To help make sense of the sensibility of the slopes I've added the average weight to the weights so we're back on the normal scale and not the mean centered scale and I've added two dashed lines to indicate the minimum and maximum height we can expect.

```{r}
#| label: prior-predictive-check-weight
#| fig-cap: A prior predictive check of the relationship between weight and height
draws <- spread_draws(
  model_height_weight_prior, b_Intercept, b_weight_mc
)

weight_mean <- data %>%
  pull(weight) %>%
  mean()

ggplot(data, aes(x = weight_mc, y = height)) +
  geom_blank() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = 272, linetype = "dashed") +
  geom_abline(
    data = filter(draws, .draw <= 100),
    mapping = aes(intercept = b_Intercept, slope = b_weight_mc),
    alpha = .25
  ) +
  geom_label(x = 0, y = 260, label = "Tallest person ever") +
  labs(x = "Weight", y = "Height") +
  scale_x_continuous(labels = function(x) round(x + weight_mean))
```

The plot shows a wide range of possible slopes, some of which are definitely unlikely. We should lower our uncertainty by reducing the standard deviation on the prior. In the next model I lower it to 3. Additionally, the negative slopes are all pretty unlikely because we should expect a positive relationship between weight and height (taller people tend to be heavier). We could therefore also change our prior to force it to be positive using the `lb` argument in our prior for `b`. Let's not do this though. Let's assume we have no idea whether the relationship will be positive or negative and instead focus on the standard deviation instead so that we don't obtain relationships we definitely know are unlikely (e.g., if many of the lines allow for many tallest people ever or heights below zero).

```{r}
#| label: weight-updated-prior
model_height_weight_prior <- brm(
  height ~ weight_mc,  
  data = data, 
  family = gaussian,
  prior = c(
      prior(normal(170, 10), class = "Intercept"),
      prior(cauchy(5, 5), class = "sigma"),
      prior(normal(0, 3), class = "b")
    ), 
  cores = 4,
  seed = 4, 
  sample_prior = "only", 
  file = "models/model_height_weight_prior_3.rds"
)
```

Let's inspect the lines again.

```{r}
#| label: prior-predictive-check-weight-update
#| fig-cap: A prior predictive check of the relationship between weight and height
draws <- spread_draws(
  model_height_weight_prior, b_Intercept, b_weight_mc, ndraws = 100
)

weight_mean <- data %>%
  pull(weight) %>%
  mean()

ggplot(data, aes(x = weight, y = height)) +
  geom_blank() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = 272, linetype = "dashed") +
  geom_abline(
    data = draws,
    mapping = aes(intercept = b_Intercept, slope = b_weight_mc),
    alpha = .25
  ) +
  geom_label(x = 0, y = 260, label = "Tallest person ever") +
  labs(x = "Weight", y = "Height") #+
  #scale_x_continuous(labels = function(x) round(x + weight_mean))
```

This looks a lot better, so let's run the model for real now.

```{r}
#| label: weight-model
model_height_weight <- brm(
  data = data, 
  height ~ weight_mc,
  family = gaussian,
  prior = c(
      prior(normal(170, 10), class = "Intercept"),
      prior(cauchy(5, 5), class = "sigma"),
      prior(normal(0, 3), class = "b", lb = 0)
    ), 
  cores = 4,
  seed = 4,
  sample_prior = TRUE,
  file = "models/model_height_weight.rds"
)

model_height_weight
```

We see that the estimate for the weight predictor is `r round(fixef(model_height_weight)[2, 1], 2)`. Let's plot the entire posterior for the slope estimate and also compare it to the prior we set for it.

```{r}
#| label: weight-prior-posterior
#| warning: false
results <- model_height_weight %>%
  gather_draws(prior_b, b_weight_mc) %>%
  mutate(
    distribution = if_else(
      str_detect(.variable, "prior"), "prior", "posterior"
    )
  )

ggplot(results, aes(x = .value, fill = fct_rev(distribution))) +
  geom_histogram(binwidth = 0.05, position = "identity", alpha = .85) +
  xlim(0, 5) +
  labs(x = "Slope", y = "", fill = "Distribution") +
  scale_fill_viridis(option = "mako", discrete = TRUE, begin = .25, end = .75)
```

Apparently our prior was still very uninformed because the posterior shows we can be confident in a much narrower range of slopes!

Let's also create another plot in which we plot the model results against the observed data. In the graph below we plot the raw data as well as 100 lines drawn from the posterior.

```{r}
slopes_qi <- tibble(
    weight_mc = seq(from = -20, to = 20, by = 1)
  ) %>%
  add_epred_draws(model_height_weight) %>%
  median_qi()

ggplot() +
  geom_ribbon(
    mapping = aes(ymin = .lower, ymax = .upper, x = weight_mc),
    data = slopes_qi,
    alpha = .25
  ) +
  geom_line(
    mapping = aes(x = weight_mc, y = .epred),
    data = slopes_qi
  ) + 
  geom_point(
    mapping = aes(x = weight_mc, y = height),
    data = data,
    alpha = .25
  ) +
  scale_x_continuous(labels = function(x) round(x + weight_mean))
```

```{r}
predicted_slopes_qi <- tibble(
    weight_mc = seq(from = -20, to = 20, by = 1)
  ) %>%
  add_predicted_draws(model_height_weight) %>%
  median_qi()

ggplot() +
  geom_ribbon(
    aes(ymin = .lower, ymax = .upper, x = weight_mc),
    data = predicted_slopes_qi,
    alpha = .25
  ) +
  geom_line(
    aes(x = weight_mc, y = .prediction),
    data = predicted_slopes_qi
  ) + 
  geom_point(
    aes(x = weight_mc, y = height),
    data = data,
    alpha = .25
  ) +
  scale_x_continuous(labels = function(x) round(x + weight_mean))
```

### Thinking correlations instead

Maybe one reason our prior was so uninformed was because it's harder to think of the right prior for a content-specific topic such as weights and heights of the !Kung San. Maybe we can instead standardize both the heights and weights in order to turn the regression model into a simple correlation analysis. That way we can specify a prior on what we think the correlation should be, which may be easier to do because we then think in terms of whether we think the relationship is small or medium or large, or something along those lines.

So, let's standardize the heights and weights.

```{r}
#| label: standardize-heights-weights
data <- mutate(
  data, 
  height_z = (height - mean(height)) / sd(height),
  weight_z = (weight - mean(weight)) / sd(weight)
)
```

The formula for our correlation analysis is `height_z ~ weight_z`. Which priors we have to specify remains the same, but what these priors should be did change. For instance, we know that the Intercept has to be 0 now because the heights have been standardized. This means the mean will be 0. In `brms`, we can specify a constant as a prior using `constant()`.

What should the prior for $\sigma$ be? Let's keep that one the same for now and see what the posterior results will be.

The prior for the slope is a lot easier now. We can simply specify a normal distribution with a mean of 0 and a standard deviation equal to the size of the effect we deem likely, together with a lower bound of 0 and upper bound of 1.

```{r}
#| label: model-height-weight_z
model_height_weight_z <- brm(
  height_z ~ weight_z,  
  data = data, 
  family = gaussian,
  prior = c(
      prior(constant(0), class = "Intercept"),
      prior(cauchy(0, 1), class = "sigma", ub = 1),
      prior(normal(0, 1), class = "b", lb = 0, ub = 1)
    ), 
  cores = 4,
  seed = 4, 
  sample_prior = TRUE,
  file = "models/model_height_weight_prior_z.rds", file_refit = "always"
)

model_height_weight_z
```
